---
title: "Collated Bioinformatics Findings on Most Abundant Influenza Defective Genomes"
format: html
editor: visual
---

## Length Analysis of Abundant Influenza Defective Genomes Across Patient Samples

## Data and Reproducibility Notes

As portions of the data used in this project are part of ongoing research, all data import (read-in) code has been removed from this repository to maintain academic integrity and protect unpublished results. With this exception, all code presented here is complete and functional, provided the corresponding data files are supplied.

Findings related to patient hospitalisation status have also been omitted, as this data is not yet publicly available. All other findings presented can be reproduced using publicly available data.

## Project Background

Defective viral genomes (DVGs) are truncated forms of a wildtype virus that have lost a large, functionally important portion of their genome. This loss typically renders them incapable of independent replication; however, DVGs can persist and proliferate by exploiting the replicative machinery (the "public goods") produced by wildtype virus during coinfection. In doing so, DVGs can interfere with wildtype replication and, in some cases, outcompete it. This makes DVGs of considerable scientific interest: because they are replication-incompetent on their own and depend entirely on coinfection with a wildtype virus, they represent a potential avenue for future antiviral therapies. This project investigates the length and positional characteristics of abundant DVGs across patient influenza samples, with the aim of establishing a principled bin length for grouping DVGs by segment, a necessary foundation for any subsequent analysis of DVG diversity and abundance.

## Overview

This repository presents the main finding from my summer project, followed by the supplementary and preliminary analyses that informed and led to it. Each section includes the associated analysis code.

## Determining an Optimal Bin Length for Defective Viral Genome (DVG) Grouping

### Motivation

Communities of DVGs vary substantially in both the position and size of individual genomes within a given segment. Without first characterising the extent of this length variation, prior analyses lacked a principled basis for grouping DVGs, and so did not produce conclusive or meaningful results. The aim of this analysis was therefore to determine the bin length at which the greatest number of DVGs within a segment would be correctly grouped as a single DVG, rather than being split into multiple distinct DVGs.

### Method

I developed a function that takes a dataset and a segment name as parameters and performs the following steps:

1.  Identifies the most abundant DVG (by total read count) present in each sample for the given segment.

2.  Performs pairwise comparisons between DVGs to calculate the absolute difference in length, accounting for the genomic position of each DVG.

3.  Iterates over a range of cutoff values, from 1 up to the largest absolute difference observed in the dataset. For each cutoff, every pairwise absolute difference is classified as `TRUE` (≤ cutoff) or `FALSE` (\> cutoff). The loop continues until all values are classified as `TRUE`, which necessarily occurs once the cutoff reaches the maximum observed difference.

4.  For each cutoff value, calculates the proportion of `TRUE` classifications, producing a cumulative proportion curve that reaches 1 at the maximum cutoff.

This function was applied independently to each genomic segment, and the results were visualised as a faceted line plot of cumulative proportion (of `TRUE` values) against cutoff value, faceted by segment.

```{r}
#| label: Loading in Libraries for this Report

library(tidyverse)
library(dplyr)
library(tidyr)
library(magrittr)
library(vegan)
library(ggplot2)
library(purrr)
library(patchwork)
library(forcats)
library(FactoMineR)
library(factoextra)
library(gplots)
library(corrplot)
library(RColorBrewer)
library(ggbeeswarm)
library(plotly)
library(pacman)
library(usedist)
library(betareg)
library(statmod)
library(performance)
library(car)
library(broom)
library(emmeans)
```

```{r}
#| label: Comparing Absolute Difference of Most Abundant Cheats in Each Segment for Each Sample

#Function which has nested functions for comparing and storing pairwise absolute difference values, then runs a loop to find the proportion of pairwise comparisons which are below or equal to a certain threshhold (cutoff) absolute difference, and then the proportion of true is found for each cutoff point and stored in a two column dataframe with one column holding the cutoff size and the other holding the proportion true
process_segment <- function(segment_name, full_dataset) {
  
  # Filter the dataset to keep only rows belonging to the selected segment
  filtered_df <- full_dataset |> 
    filter(Segment == segment_name) 
  
  # For each SRR, keep the row with the highest TotalRC value
  ma <- filtered_df |> 
    group_by(SRR) |> 
    slice_max(order_by = TotalRC, n = 1, with_ties = FALSE) |> 
    as.data.frame()
  
  # Use the SRR values as row names
  rownames(ma) <- ma$SRR
  
  # Define a helper function to calculate pairwise differences
  pairwise_diff <- function(df){
    
    # Number of rows (one row per SRR)
    n_cols <- nrow(df)
    
    # Create a square matrix initialized with 1s
    # Rows and columns represent SRRs
    diffmatrix <- matrix(1, nrow = n_cols, ncol = n_cols)
    
    # Label the rows and columns with SRR names
    rownames(diffmatrix) <- rownames(df)
    colnames(diffmatrix) <- rownames(df)
    
    # Extract the End and Start coordinates
    end_col <- df$End
    start_col <- df$Start
    
    # Compare every pair of SRRs
    for (i in 1:(n_cols - 1)) {
      
      # Compare row i with every following row
      for (j in (i + 1):n_cols) {
        
        # Coordinates for sample i
        start_i <- start_col[i]
        end_i <- end_col[i]
        
        # Coordinates for sample j
        start_j <- start_col[j]
        end_j <- end_col[j]
        
        # Calculate the total absolute difference in start and end positions
        absdiff <- abs(start_i - start_j) + abs(end_i - end_j)
        
        # Store the value in the upper triangle
        diffmatrix[i, j] <- absdiff
        
        # Store the same value in the lower triangle (matrix is symmetric)
        diffmatrix[j, i] <- absdiff
      }
    }
    
    # Return the completed distance matrix
    return(diffmatrix)
  }
  
  # Convert the pairwise difference matrix into a long-format table
  ma_df <- pairwise_diff(ma) |> 
    
    # Convert matrix to data frame
    as.data.frame() |> 
    
    # Move row names into a column called SRR
    rownames_to_column(var = "SRR") |> 
    
    # Convert from wide format to long format
    # Each row becomes one SRR comparison
    pivot_longer(cols = -SRR, names_to = "SRR_Pair") |> 
    
    # Create a combined identifier for each SRR pair
    unite("Pairs", SRR, SRR_Pair, sep = "_", remove = FALSE) |> 
    
    # Keep only the pair identifier and difference value
    select(Pairs, value) |> 
    
    # Remove self-comparisons and identical coordinates (difference ≤ 1)
    filter(value > 1) |> 
    
    # Convert to a standard data frame
    as.data.frame()
  
  # Define a helper function to create cumulative cutoff tables
  cutoff <- function(df) {
    
    # Extract the difference values
    diff <- df[[2]]
    
    # Find the largest observed difference
    max_diff <- max(diff, na.rm = TRUE)
    
    # Create an empty matrix
    # Rows = SRR pairs
    # Columns = cutoff values from 1 to max_diff
    results_matrix <- matrix(
      NA,
      nrow = length(diff),
      ncol = max_diff
    )
    
    # For each cutoff value
    for (bin in 1:max_diff) {
      
      # Record TRUE if the pairwise difference is less than or equal to the cutoff
      results_matrix[, bin] <- diff <= bin
    }
    
    # Convert matrix to data frame
    results_df <- as.data.frame(results_matrix)
    
    # Name columns according to the cutoff values
    colnames(results_df) <- paste0(1:max_diff)
    
    # Use pair names as row names
    rownames(results_df) <- df[[1]]
    
    # Return the logical cutoff table
    return(results_df)
  }
  
  # Calculate the cumulative proportion of pairs below each cutoff
  prop_diff <- cutoff(ma_df) |> 
    
    # Transpose so each row represents one cutoff value
    t() |> 
    
    # Convert to data frame
    as.data.frame() |> 
    
    # Calculate the proportion of TRUE values for each cutoff
    mutate(proportion_true = rowMeans(pick(where(is.logical)), na.rm = TRUE)) |> 
    
    # Move row names (cutoff values) into a column
    rownames_to_column(var = "bin_size") |>
    
    # Convert cutoff labels to numeric values
    mutate(bin_numeric = as.numeric(bin_size)) |> 
    
    # Keep only the cutoff and its corresponding proportion
    select(bin_numeric, proportion_true)
  
  # Return the final table of cumulative proportions
  return(prop_diff)
}

#List of all of the influenza segments so that the above function can be run for all
segments <- c('PB1', 'PB2', 'PA', 'NA', 'HA', 'NP', 'MP', 'NS')

#Stores a list of dataframes for each segment which has the above function applied to it using the full dataset 'merged'
results_list <- lapply(segments, process_segment, full_dataset = merged)
names(results_list) <- segments
combined_df <- bind_rows(results_list, .id = "Segment")

#Finding the number of samples per segment
sample_counts <- merged %>%
  distinct(Segment, SRR) %>%
  count(Segment)%>%
  mutate(label = paste0("n = ", n))

#Rowan colour palette
colors <- c(
  "PB2" = "#DD4124",
  "PB1" = "#ED8B00",
  "PA" = "#EDD746",
  "HA" = "#003552",
  "NP" = "#0F85A0",
  "NA" = "#A4D0F4",
  "MP" = "#61599D",
  "NS" = "#EBAFAD"
)

#Facetted plot
ggplot(combined_df, aes(x = bin_numeric, y = proportion_true, color = Segment)) +
  geom_line() +
  facet_wrap(~Segment, ncol = 4, scales = "free_x") + 
  theme_minimal()+
  scale_color_manual(values = colors) +
  geom_text(
    data = sample_counts,
    aes(label = label),
    x = Inf, y = -Inf,         
    hjust = 1.0, vjust = -2.0,  
    inherit.aes = FALSE        
  ) +
  labs(
    title = "Proportion of Pairs Within Distance Cutoff by Segment",
    x = "Distance Bin Cutoff",
    y = "Proportion (True)",
    color = "Comparison Group"
  )+
  theme(legend.position = "none")

```

### Findings and Follow-Up

Discussion of these results with my supervisors highlighted a distinct inflection ("kink") in the resulting curves, most notably in the polymerase segments. I proposed that this pattern reflects the presence of two positionally distinct subpopulations of DVGs within a segment, which converge into a single group only at large cutoff values, owing to their substantial difference in genomic position.

To test this interpretation, I repeated the analysis using only the absolute length difference between DVGs, excluding positional information. We further discussed extending this analysis to patient groups stratified by hospitalisation status, to assess whether the trend and inflection point differed within or between groups. We hypothesised that hospitalised individuals would show earlier plateauing, reflecting more homogeneous DVG communities. As this component is part of ongoing research, the corresponding results are not reported here; however, the code for both analyses (grouped and length-only) is provided below.

The length-only analysis did not reproduce the prominent inflection observed previously, supporting the interpretation that DVG position, rather than length alone, underlies the pattern, and lending further support to the presence of two distinct DVG subtypes within these samples.

To investigate this further, I adapted a visualisation approach provided by my supervisor, in which each DVG is plotted as a line representing its length and start/end position relative to other DVGs. I then plotted DVG end-tail position against segment length, coloured by DVG load, to assess whether larger deletions were associated with higher overall DVG load. This did not yield a conclusive result, as no clear pattern emerged.

I also conducted a deeper analysis of samples whose most abundant cheat was shorter than 1000 bp, subsampling these from the full dataframe and visualising their five most abundant cheats. This was done to investigate whether the most abundant DVG might be in linkage with a larger DVG. The results showed that many of these samples also contained a large DVG present at high abundance; however, their start and end sites differed by a minimum of 15 nucleotides.

More detailed linkage analysis showed that of the 34 samples analysed, 14 had no detectable linkage events, while the remaining samples had between 1 and 20 linked events. The variation in the number of linked events suggests that small deletions may accumulate over time and potentially contribute to the formation of larger deletions, although multiple independent small deletions may also occur within the same sample.

I also compared DVG abundance between samples with and without detectable linkage events. Samples with linkage events showed considerably higher cheat loads than samples without detected linkage, suggesting a potential association between deletion linkage and DVG abundance.

The code below outlines the construction of the plots used for each of the analyses discussed above.

```{r}
#| label: Same as Before but Using Absolute Length Difference Only

process_length <- function(segment_name, full_dataset) {
  
  # Filter the dataset to keep only rows belonging to the selected segment
  filtered_df <- full_dataset |> 
    filter(Segment == segment_name) 
  
  # For each SRR, keep the row with the highest TotalRC value
  ma <- filtered_df |> 
    group_by(SRR) |> 
    slice_max(order_by = TotalRC, n = 1, with_ties = FALSE) |> 
    as.data.frame()
  
  # Use the SRR values as row names
  rownames(ma) <- ma$SRR
  
  # Define a helper function to calculate pairwise differences
  pairwise_diff <- function(df){
    n_cols <- nrow(df)
    diffmatrix <- matrix(1, nrow = n_cols, ncol = n_cols)
    rownames(diffmatrix) <- rownames(df)
    colnames(diffmatrix) <- rownames(df)
    for (i in 1:(n_cols - 1)) {
      for (j in (i + 1):n_cols) {
        absdiff <- abs(df$Length[i]) - abs(df$Length[j]) #This is the line of interest which differs from previous block for this analysis
        diffmatrix[i, j] <- absdiff
        diffmatrix[j, i] <- absdiff
      }
    }
    return(diffmatrix)
  }
  
  # Convert the pairwise difference matrix into a long-format table
  ma_df <- pairwise_diff(ma) |> 
    
    # Convert matrix to data frame
    as.data.frame() |> 
    
    # Move row names into a column called SRR
    rownames_to_column(var = "SRR") |> 
    
    # Convert from wide format to long format
    # Each row becomes one SRR comparison
    pivot_longer(cols = -SRR, names_to = "SRR_Pair") |> 
    
    # Create a combined identifier for each SRR pair
    unite("Pairs", SRR, SRR_Pair, sep = "_", remove = FALSE) |> 
    
    # Keep only the pair identifier and difference value
    select(Pairs, value) |> 
    
    # Remove self-comparisons and identical coordinates (difference ≤ 1)
    filter(value > 1) |> 
    
    # Convert to a standard data frame
    as.data.frame()
  
  # Define a helper function to create cumulative cutoff tables
  cutoff <- function(df) {
    
    # Extract the difference values
    diff <- df[[2]]
    
    # Find the largest observed difference
    max_diff <- max(diff, na.rm = TRUE)
    
    # Create an empty matrix
    # Rows = SRR pairs
    # Columns = cutoff values from 1 to max_diff
    results_matrix <- matrix(
      NA,
      nrow = length(diff),
      ncol = max_diff
    )
    
    # For each cutoff value
    for (bin in 1:max_diff) {
      
      # Record TRUE if the pairwise difference is less than or equal to the cutoff
      results_matrix[, bin] <- diff <= bin
    }
    
    # Convert matrix to data frame
    results_df <- as.data.frame(results_matrix)
    
    # Name columns according to the cutoff values
    colnames(results_df) <- paste0(1:max_diff)
    
    # Use pair names as row names
    rownames(results_df) <- df[[1]]
    
    # Return the logical cutoff table
    return(results_df)
  }
  
  # Calculate the cumulative proportion of pairs below each cutoff
  prop_diff <- cutoff(ma_df) |> 
    
    # Transpose so each row represents one cutoff value
    t() |> 
    
    # Convert to data frame
    as.data.frame() |> 
    
    # Calculate the proportion of TRUE values for each cutoff
    mutate(proportion_true = rowMeans(pick(where(is.logical)), na.rm = TRUE)) |> 
    
    # Move row names (cutoff values) into a column
    rownames_to_column(var = "bin_size") |>
    
    # Convert cutoff labels to numeric values
    mutate(bin_numeric = as.numeric(bin_size)) |> 
    
    # Keep only the cutoff and its corresponding proportion
    select(bin_numeric, proportion_true)
  
  # Return the final table of cumulative proportions
  return(prop_diff)
}

#List of all of the influenza segments so that the above function can be run for all
segments <- c('PB1', 'PB2', 'PA', 'NA', 'HA', 'NP', 'MP', 'NS')

#Stores a list of dataframes for each segment which has the above function applied to it using the full dataset 'merged'
results_list <- lapply(segments, process_length, full_dataset = merged)
names(results_list) <- segments
combined_df <- bind_rows(results_list, .id = "Segment")

#Finding the number of samples per segment
sample_counts <- merged %>%
  distinct(Segment, SRR) %>%
  count(Segment)%>%
  mutate(label = paste0("n = ", n))

#Rowan colours
colors <- c(
  "PB2" = "#DD4124",
  "PB1" = "#ED8B00",
  "PA" = "#EDD746",
  "HA" = "#003552",
  "NP" = "#0F85A0",
  "NA" = "#A4D0F4",
  "MP" = "#61599D",
  "NS" = "#EBAFAD"
)

#Facetted plot
ggplot(combined_df, aes(x = bin_numeric, y = proportion_true, color = Segment)) +
  geom_line() +
  facet_wrap(~Segment, ncol = 4, scales = "free_x") + 
  theme_minimal()+
  scale_color_manual(values = colors) +
  geom_text(
    data = sample_counts,
    aes(label = label),
    x = Inf, y = -Inf,         
    hjust = 1.0, vjust = -2.0,  
    inherit.aes = FALSE        
  ) +
  labs(
    title = "Proportion of Pairs Within Distance Cutoff by Segment (Length Only)",
    x = "Distance Bin Cutoff",
    y = "Proportion (True)",
    color = "Comparison Group"
  )+
  theme(legend.position = "none")

```

```{r}
#| label: Community Comparison of Absolute Difference in Most Abundant Cheats 

process_community <- function(segment_name, full_dataset) {
  
  # Keep only rows corresponding to the selected genome segment
  filtered_df <- full_dataset |> 
    filter(Segment == segment_name) 
  
  # For each sample (SRR), retain only the most abundant deletion (highest TotalRC)
  ma <- filtered_df |> 
    group_by(SRR) |> 
    slice_max(order_by = TotalRC, n = 1, with_ties = FALSE) |> 
    as.data.frame()
  
  # Set the row names to the SRR identifiers for easier indexing later
  rownames(ma) <- ma$SRR
  
  # Function to calculate pairwise distances between all samples
  pairwise_diff <- function(df){
    
    # Number of samples (rows) in the dataframe
    n_cols <- nrow(df)
    
    # Create an empty square matrix to store pairwise distances
    # Initialise with 1s so the diagonal is not zero (helps remove self-comparisons later)
    diffmatrix <- matrix(1, nrow = n_cols, ncol = n_cols)
    
    # Assign row and column names using the SRR identifiers
    rownames(diffmatrix) <- rownames(df)
    colnames(diffmatrix) <- rownames(df)
    
    # Extract deletion end coordinates
    end_col <- df$End
    
    # Extract deletion start coordinates
    start_col <- df$Start
    
    # Compare every sample against every other sample
    for (i in 1:(n_cols - 1)) {
      for (j in (i + 1):n_cols) {
        
        # Calculate the pairwise distance as the sum of the absolute
        # differences between the start and end coordinates
        absdiff <- abs(start_col[i] - start_col[j]) + abs(end_col[i] - end_col[j])
        
        # Store the distance in the upper triangle of the matrix
        diffmatrix[i, j] <- absdiff
        
        # Mirror the value into the lower triangle to keep the matrix symmetric
        diffmatrix[j, i] <- absdiff
      }
    }
    
    # Return the completed pairwise distance matrix
    return(diffmatrix)
  }
  
  # Convert the pairwise distance matrix into a long-format dataframe
  ma_df <- pairwise_diff(ma) |> 
    
    # Convert the matrix to a dataframe
    as.data.frame() |> 
    
    # Move row names (SRR IDs) into a column
    rownames_to_column(var = "SRR") |> 
    
    # Convert from wide format to long format (one row per sample pair)
    pivot_longer(cols = -SRR, names_to = "SRR_Pair") |> 
    
    # Create a unique identifier for each pair of samples
    unite("Pairs", SRR, SRR_Pair, sep = "_", remove = FALSE) |> 
    
    # Keep only the relevant columns
    select(Pairs, SRR, SRR_Pair, value) |> 
    
    # Remove self-comparisons (distance = 1 from initialisation)
    filter(value > 1) |> 
    
    # Add hospitalization status for the first sample
    left_join(Meta, by = c("SRR" = "Run")) |> 
    rename(Status_1 = `Hospitalized?`) |> 
    
    # Add hospitalization status for the second sample
    left_join(Meta, by = c("SRR_Pair" = "Run")) |> 
    rename(Status_2 = `Hospitalized?`) |> 
    
    # Create a comparison group (e.g. No_No, No_Yes, Yes_Yes)
    # pmin/pmax ensure that No_Yes and Yes_No are treated as the same group
    mutate(Group = paste(pmin(Status_1, Status_2),
                         pmax(Status_1, Status_2),
                         sep = "_")) |> 
    
    # Keep only the pairwise distance and comparison group
    select(value, Group)
  
  # Find the largest pairwise distance, ignoring any missing values
  max_diff <- max(ma_df$value, na.rm = TRUE)
  
  # Generate a sequence of distance thresholds from 1 up to the maximum distance
  bin_sequence <- 1:max_diff
  
  # For each distance threshold, calculate the proportion of sample pairs
  # that fall within that cutoff for each hospitalization group
  prop_diff <- purrr::map_dfr(bin_sequence, function(bin) {
    ma_df |>
      
      # Analyse each hospitalization group separately
      group_by(Group) |>
      
      # Calculate the proportion of distances less than or equal to the cutoff
      summarise(
        proportion_true = mean(value <= bin, na.rm = TRUE),
        .groups = "drop"
      ) |>
      
      # Record the current cutoff value for plotting
      mutate(bin_numeric = bin)
  })
  
  # Return the summary dataframe for this genome segment
  return(prop_diff)
}

sample_counts <- merged %>%
  distinct(Segment, SRR) %>%
  count(Segment)%>%
  mutate(label = paste0("n = ", n))

segments <- c('PB1', 'PB2', 'PA', 'NA', 'HA', 'NP', 'MP', 'NS')

#Stores a list of dataframes for each segment which has the above function applied to it using the full dataset 'merged'
results_list <- lapply(segments, process_community, full_dataset = merged)
names(results_list) <- segments
combined_df <- bind_rows(results_list, .id = "Segment")

ggplot(combined_df, aes(x = bin_numeric, y = proportion_true, color = Group, group = Group)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Segment, ncol = 4, scales = "free_x") + 
  theme_minimal() +
  scale_color_manual(values = c(
    "Yes_Yes" = "#D55E00", 
    "No_Yes"  = "#E69F00", 
    "No_No"   = "#009E73"
  )) +
  geom_text(
    data = sample_counts,
    aes(label = label),
    x = Inf, y = -Inf,         
    hjust = 1.0, vjust = -2.0,  
    inherit.aes = FALSE        
  ) +
  labs(
    title = "Proportion of Pairs Within Distance Cutoff by Segment, Grouped by Hospitalisation",
    x = "Distance Bin Cutoff",
    y = "Proportion (True)",
    color = "Comparison Group"
  ) +
  theme(legend.position = "bottom") 
```

```{r}
#| label: Visualising the Cheat Deletions

visualise_prep <- function(segment_name, full_dataset) {
  
  # Filter the dataset to keep only rows belonging to the selected segment
  filtered_df <- full_dataset |> 
    filter(Segment == segment_name) 
  
  # For each SRR, keep the row with the highest TotalRC value
  ma <- filtered_df |> 
    group_by(SRR) |> 
    slice_max(order_by = TotalRC, n = 1, with_ties = FALSE) |> 
    as.data.frame()
  
  #Adds a Number to Each Cheat so that they can be Ordered
  ma$Unique <- seq_len(nrow(ma))
  
  return(ma)
}

segments <- c('PB1', 'PB2', 'PA', 'NA', 'HA', 'NP', 'MP', 'NS')

#Stores a list of dataframes for each segment which has the above function applied to it using the full dataset 'merged'
results_list <- lapply(segments, visualise_prep, full_dataset = DVGs)
names(results_list) <- segments
combined_df <- bind_rows(results_list, .id = "Segment")

combined_df <- combined_df %>%
  
  # Group the data by genome segment so numbering restarts within each segment
  group_by(Segment) %>%
  
  # Sort deletions by their start coordinate within each segment
  arrange(Length, .by_group = TRUE) %>%
  
  # Assign each deletion a unique sequential number within its segment
  # (e.g. PB1: 1, 2, 3...; PB2: 1, 2, 3...)
  mutate(Unique = row_number()) %>%
  
  # Remove the grouping so later operations treat the dataframe as a whole
  ungroup()
  
p4<- ggplot(combined_df) +
  facet_wrap(~Segment, ncol = 4, scales = "free") +
  geom_segment(aes(x = Start,
                   xend = End,
                   y = Unique,
                   yend = Unique,
                   colour = 'red'),
               linewidth = 0.2,
               alpha = 1)  +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        legend.position = "none")
```

```{r}
#| label: Cheat load Comparison Between Most Abundant Cheats

#Running the exact same function as previously, so did not include in this chunk
results_list <- lapply(segments, visualise_prep, full_dataset = cheatload)
names(results_list) <- segments
combined_df <- bind_rows(results_list, .id = "Segment")

#Process the combined dataframe to create unique identifiers and calculate deletion sizes
combined_df <- combined_df %>%
  
  #Perform the following operations separately for each influenza genome segment
  group_by(Segment) %>%
  
  #Sort deletions by their start coordinate within each segment
  arrange(Start, .by_group = TRUE) %>%
  
  #Assign a unique sequential number to each deletion within a segment
  mutate(Unique = row_number()) %>%
  
  #Remove the grouping so subsequent operations act on the entire dataframe
  ungroup() %>%
  
  #Calculate the size of each deletion (end position minus start position)
  mutate(Deletion_Size = End - Start) %>%
  
  #Sort the data first by genome segment and then by deletion size
  arrange(Segment, Deletion_Size) %>%
  
  #Create identifiers for plotting and preserve their order
  mutate(
    
    #Combine the segment name and unique number into a unique deletion ID
    Unique_ID = paste0(Segment, "_", Unique),
    
    #Convert the IDs to an ordered factor so plots retain this ordering
    Unique_Ordered = fct_inorder(Unique_ID)
  )

ggplot(combined_df) +
  facet_wrap(~Segment, ncol = 4, scales = "free") +
  geom_segment(aes(x = 1, 
                   xend = Length, 
                   y = Unique_Ordered, 
                   yend = Unique_Ordered,
                   colour = DVG_abundance_focal_local_random),
               linewidth = 0.5,
               alpha = 0.7) +
  geom_segment(aes(x = Start, 
                   xend = End, 
                   y = Unique_Ordered, 
                   yend = Unique_Ordered), 
               colour = "red",
               linewidth = 0.8,
               alpha = 0.9) +
  scale_colour_viridis_c(option = "plasma", name = "Cheat Load") + 
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

```{r}
#| label: Analysis of Samples with Short Dominant DVGs

# Function to identify samples where the most abundant DVG is shorter than 1000 bp
smallcheats <- function(segment_name, df) {
  
  df |>
    # Select data for the specified viral segment
    filter(Segment == segment_name) |>
    
    # Group the data by sample
    group_by(SRR) |>
    
    # Identify the single most abundant DVG in each sample
    slice_max(
      order_by = TotalRC,
      n = 1,
      with_ties = FALSE
    ) |>
    
    # Retain samples where the most abundant DVG is shorter than 1000 bp
    filter(DelLength < 1000) |>
    
    # Extract the sample IDs
    pull(SRR)
}

# Define the viral segments to analyse
segments <- c("PB1", "PB2", "PA")

# Apply the smallcheats function to each viral segment
results_list <- lapply(
  segments,
  smallcheats,
  df = cheatload
)

# Assign the segment names to the resulting list
names(results_list) <- segments

# Combine the selected sample IDs and their corresponding segments into a data frame
selected_srrs <- data.frame(
  Segment = rep(names(results_list), lengths(results_list)),
  SRR = unlist(results_list)
)

# Select the top five most abundant DVGs from each selected sample
plot_data <- cheatload |>
  
  # Retain only samples identified as having a short dominant DVG
  inner_join(
    selected_srrs,
    by = c("Segment", "SRR")
  ) |>
  
  # Group the data by viral segment and sample
  group_by(Segment, SRR) |>
  
  # Select the five most abundant DVGs in each sample
  slice_max(
    order_by = TotalRC,
    n = 5,
    with_ties = FALSE
  ) |>
  
  # Order the DVGs from highest to lowest abundance within each sample
  arrange(desc(TotalRC), .by_group = TRUE) |>
  
  # Assign an order number to each DVG for plotting
  mutate(
    Deletion_order = row_number()
  ) |>
  
  # Remove the grouping structure
  ungroup()

# Function to generate plots for an individual viral segment
plot_segment <- function(segment_name, data) {
  
  data |>
    # Select data for the specified viral segment
    filter(Segment == segment_name) |>
    
    # Initialise the ggplot
    ggplot() +
    
    # Create a separate panel for each sample
    facet_wrap(
      ~SRR,
      ncol = 4,
      scales = "free"
    ) +
    
    # Draw a line representing the full viral segment for each DVG
    geom_segment(
      aes(
        x = 1,
        xend = Length,
        y = Deletion_order,
        yend = Deletion_order,
        # Colour the segment according to hospitalisation status
        colour = `Hospitalized?`,
        # Alternative colouring by DVG abundance
        # colour = DVG_abundance_focal_local_random,
      ),
      linewidth = 0.8,
      alpha = 0.9
    ) +
    
    # Overlay the deleted region of each DVG in red
    geom_segment(
      aes(
        x = Start,
        xend = End,
        y = Deletion_order,
        yend = Deletion_order
      ),
      colour = "red",
      linewidth = 1.2,
      alpha = 1.3
    ) +
    
    # Alternative continuous colour scale based on DVG abundance
    # scale_colour_viridis_c(
    #   option = "plasma",
    #   name = "Cheat Load"
    # ) +
    
    # Set a discrete colour scale based on hospitalisation status
    scale_colour_viridis_d(
      option = "plasma",
      name = "Hospitalisation"
    ) +
    
    # Apply a clean, classic ggplot theme
    theme_classic() +
    
    # Remove y-axis labels and tick marks
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
}

# Generate the plot for the PB1 segment
pb1_plot <- plot_segment("PB1", plot_data)

# Generate the plot for the PB2 segment
pb2_plot <- plot_segment("PB2", plot_data)

# Generate the plot for the PA segment
pa_plot <- plot_segment("PA", plot_data)

```

```{r}
#| label: Small-event linkage analysis and abundance comparison

# Find the most abundant small deletion for each sample in each segment
smallcheats <- function(segment_name, df2) {
  df2 |>
    filter(Segment == segment_name) |>
    group_by(Sample) |>
    slice_max(order_by = TotalRC, n = 1, with_ties = FALSE) |>
    filter(DelLength < 1000) |>
    select(Sample, Start, End)
}

segments <- c("PB1", "PB2", "PA")

results_list <- lapply(segments, smallcheats, df2 = cheatload)
names(results_list) <- segments

small_samples <- bind_rows(results_list, .id = "Segment") |> 
  as.data.frame()


# Subset the recombination data to samples containing the small events
base <- semi_join(recombdata, small_samples, by = c("Segment", "Sample")) |> 
  filter(Type == "Del")

# Find the most frequently observed deletion event for each sample
top_events <- base |> 
  count(Start, End, Sample, RecombEvent, Segment) |> 
  group_by(Sample) |> 
  filter(n == max(n)) |> 
  ungroup()

# Keep a unique list of the selected deletion events
all_distinct <- top_events |>
  distinct(
    Sample,
    Segment,
    DeletionEvent = RecombEvent
  )


# Find other recombination events that share reads with each small event
linkage.anal <- function(df_all, df_small, segment) {
  
  df_all <- df_all |>
    filter(Segment == segment)
  
  df_small <- df_small |>
    filter(Segment == segment)
  
  results <- list()
  
  for (i in df_small$RecombEvent) {
    
    # Get the reads belonging to the current small deletion
    small_event_reads <- df_all |>
      filter(RecombEvent == i)
    
    # Find other events that share those reads
    linked_events <- df_all |>
      filter(Reads %in% small_event_reads$Reads) |>
      filter(RecombEvent != i)
    
    results[[as.character(i)]] <- linked_events
  }
  
  return(results)
}

segments <- c("PB1", "PB2", "PA")

results_list2 <- lapply(
  segments,
  linkage.anal,
  df_all = base,
  df_small = top_events
)
names(results_list2) <- segments


# Combine the linkage results from all segments and deletion events
combined_results <- imap_dfr(
  results_list2,
  function(segment_results, segment_name) {
    
    imap_dfr(
      segment_results,
      function(event_df, deletion_event) {
        
        event_df |>
          mutate(
            DeletionEvent = deletion_event
          )
      }
    )
  }
) |> 
  filter(RevCount == 2) |> 
  select(Sample, Segment, Start, End, RecombEvent, DeletionEvent) |> 
  distinct(
    Sample,
    DeletionEvent,
    RecombEvent,
    .keep_all = TRUE
  ) |> 
  group_by(Sample, Segment, DeletionEvent) |>
  mutate(
    # Number of unique start/end combinations linked to each deletion
    nunique = n_distinct(paste(Start, End))
  ) |>
  ungroup()


# Number of selected deletion events that have no detected linkage
non_linked <- n_distinct(all_distinct$Sample) - n_distinct(combined_results$Sample)


# Add the non-linked events as zeroes for the histogram only
plot_data <- combined_results |>
  group_by(Sample, Segment) |> 
  reframe(
    nunique = unique(nunique)
  ) |> 
  ungroup() |> 
  pull(nunique) |> 
  c(rep(0, non_linked))


p1 <- ggplot(data.frame(nunique = plot_data), aes(x = nunique)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5
  ) +
  labs(
    x = "Number of linked events",
    y = "Number of deletion events"
  )


# Compare DVG abundance between samples with and without linkage
linked_abun <- semi_join(abun, combined_results, by = "Sample")

all_abun <- semi_join(abun, all_distinct, by = "Sample")

combined_abun <- bind_rows(
  linked_abun |> mutate(Linked = "Yes"),
  all_abun |>  
    anti_join(linked_abun, by = "Sample") |> 
    mutate(Linked = "No")
)


p2 <- ggplot(
  combined_abun,
  aes(x = Linked, y = DVG_abundance_focal_local_random)
) +
  geom_boxplot() +
  labs(
    title = "DVG abundance based on linked or not linked small abundant cheats",
    x = "Linkage Status",
    y = "Cheat Load/DVG Abundance"
  ) +
  theme_minimal()
```

## Richness and Evenness Between Hospitalisation Groups

Using the full dataset, including every DVG present in every sample, I performed pairwise dissimilarity comparisons using the Bray–Curtis metric. Results were visualised as box plots, coloured by comparison type (within-group: Hospitalised–Hospitalised or Not Hospitalised–Not Hospitalised; between-group: Hospitalised–Not Hospitalised), followed by statistical testing to assess significance.

I additionally generated richness-versus-evenness plots to compare overall diversity across groups, and evenness-versus-DVG-abundance plots to examine whether DVG abundance was associated with DVG diversity within specific sample groups.

As this analysis is part of ongoing research, results are not reported here; the analysis pipeline is provided below.

```{r}
#| label: Calculate pairwise DVG distances by hospitalisation status

# Filter to H3N2 PB1 DVGs, bin breakpoint positions, and create a feature count matrix
H3N2_f <- filter(DVGs, Subtype == "H3N2") %>% 
  filter(Segment == "PB1") %>%
  select(cols = -Subtype) %>% 
  na.omit() %>%  # Included as a safeguard; no missing values are currently present
  mutate(
    StartBin = cut_interval(Start, length = 30, right = FALSE),
    EndBin = cut_interval(End, length = 30, right = FALSE)
  ) %>%
  mutate(feature = paste0(StartBin, "_", EndBin)) %>%
  select(cols = -c(Start, End, StartBin, EndBin, Segment)) %>%
  pivot_wider(
    names_from = feature,
    values_from = TotalRC,
    values_fill = 0,
    values_fn = sum
  ) %>%
  as.data.frame()

# Use SRR accession numbers as row names
rownames(H3N2_f) <- H3N2_f$SRR
H3N2_f <- H3N2_f[, -1]

# Merge the feature matrix with sample metadata and remove samples with unknown hospitalisation status
H3N2f_meta <- merge(H3N2_f, Meta_Data, by.x = "row.names", by.y = "SRR") %>% 
  filter(`Hospitalized?` != "Unknown")

rownames(H3N2f_meta) <- H3N2f_meta$Row.names
# 109 H3N2 samples with known hospitalisation status

# Calculate total DVG reads per sample and retain samples with at least 50 reads
rowSums(dplyr::select(H3N2f_meta, 2:(ncol(H3N2_f) + 1))) -> H3N2f_meta$Total_DVG_reads
H3N2f_meta %<>% filter(Total_DVG_reads >= 50)

# Extract only the DVG feature matrix for distance calculations
dfDist <- dplyr::select(H3N2f_meta, 2:(ncol(H3N2_f) + 1))

# Calculate Bray-Curtis distances after rarefying each sample to 50 DVG reads
dist_all <- avgdist(
  x = dfDist,
  sample = 50,
  dmethod = "bray"
)

# Calculate distance matrices separately for hospitalised and non-hospitalised samples
Hospitalised_Matrix <- filter(H3N2f_meta, `Hospitalized?` == "Yes") %>%
  dplyr::select(2:(ncol(H3N2_f) + 1)) %>%
  avgdist(dmethod = "bray", sample = 50)

Not_Matrix <- filter(H3N2f_meta, `Hospitalized?` == "No") %>%
  select(2:(ncol(H3N2_f) + 1)) %>%
  avgdist(dmethod = "bray", sample = 50)

# Remove missing values and diagonal zeros from each distance matrix
clean_not <- Not_Matrix[!is.na(Not_Matrix) & Not_Matrix != 0]
clean_hosp <- Hospitalised_Matrix[!is.na(Hospitalised_Matrix) & Hospitalised_Matrix != 0]
clean_all <- dist_all[!is.na(dist_all) & dist_all != 0]

# Convert a symmetric distance matrix into a data frame of unique sample pairs and distances
extract_pairwise_pairs <- function(mat) {
  
  mat <- as.matrix(mat)  # Ensure the input is a matrix
  
  # Identify unique pairwise comparisons using the lower triangle
  lower_indices <- lower.tri(mat)
  
  # Retrieve the sample names for each comparison
  row_names <- rownames(mat)[row(mat)[lower_indices]]
  col_names <- colnames(mat)[col(mat)[lower_indices]]
  
  # Sort sample names so each pair has a consistent label
  sorted_pairs <- pmin(row_names, col_names)
  sorted_pairs2 <- pmax(row_names, col_names)
  
  # Create unique pair identifiers
  pair_labels <- paste0(sorted_pairs, "___", sorted_pairs2)
  
  # Extract the corresponding distance values
  pair_values <- mat[lower_indices]
  
  # Return pair labels and distances as a data frame
  result_matrix <- data.frame(
    Pair = pair_labels,
    Value = pair_values
  )
  
  return(result_matrix)
}

# Convert each distance matrix into a pairwise comparison data frame
all_df <- extract_pairwise_pairs(dist_all)

nn_df <- extract_pairwise_pairs(Not_Matrix) %>%
  mutate(Type = "No_vs_No")

hh_df <- extract_pairwise_pairs(Hospitalised_Matrix) %>%
  mutate(Type = "Yes_vs_Yes")

# Identify comparisons between hospitalised and non-hospitalised samples
hospvnot_labels <- setdiff(setdiff(all_df$Pair, nn_df$Pair), hh_df$Pair)

hospvnot <- all_df$Value[all_df$Pair %in% hospvnot_labels]

hn_df <- as.data.frame(cbind(hospvnot_labels, hospvnot)) %>%
  mutate(Type = "Yes_vs_No") %>%
  rename(
    Pair = hospvnot_labels,
    Value = hospvnot
  )

# Check the number of hospitalised vs non-hospitalised comparisons
length(hospvnot)

# Combine all pairwise comparisons into a single long-format data frame
LongDf <- rbind(hn_df, nn_df, hh_df) %>%
  mutate(Value = as.numeric(Value))

```

```{r}
#| label: Boxplot and Statistical Analysis

#BoxPlot
colnames(LongDf)
str(LongDf)

ggplot(LongDf, aes(x = Type, 
                   y = Value,
                   fill = Type)) +
  geom_boxplot(outlier.shape = ".") +
  labs(title = "Distance Dissimilarity between Hospitalised and Not Hospitalised Patients in PB1 Segment Cheats", 
       x = "Distance Matrix Comparisons", 
       y = "Distance Dissimilarity (Bray-Curtis)") +
  theme_classic()


#Regression analysis and plots of residuals
test <- betareg(Value ~ Type,
                data = LongDf, link = 'logit')

check_model(test, residual_type
            = "normal")

#Statistical Tests
Anova(test) #Significant differences in bc distanbce among groups
glance(test) #no pseudo Rsq available
summary(test) #no no most disimilar followed by yes no and then yes yes
emmeans(test, pairwise~Type, adjust='fdr', mode="response") #All pairs significantly different

```

```{r}
#| label: Richness and Evenness Analysis

# Prepare H3N2 PB1 DVG data for richness and diversity analysis
H3N2_re <- filter(DVGs, Subtype == "H3N2") %>% 
  filter(Segment == "PB1") %>%
  select(cols = -Subtype) %>% 
  na.omit() %>%  # Included as a safeguard; no missing values are currently present
  mutate(
    StartBin = cut_interval(Start, length = 30, right = FALSE),
    EndBin = cut_interval(End, length = 30, right = FALSE)
  ) %>%
  mutate(feature = paste0(StartBin, "_", EndBin)) %>%
  merge(Meta_Data, by.x = "SRR", by.y = "SRR") %>%
  filter(`Hospitalized?` != "Unknown") %>% 
  select(feature, SRR, TotalRC, `Hospitalized?`)

# Calculate feature richness and Simpson diversity by hospitalisation status
Rich_Even <- H3N2_re %>%
  group_by(feature, `Hospitalized?`) %>%
  summarise(
    sobs_richness = specnumber(TotalRC),
    simpson = diversity(TotalRC, index = "simpson"),
    .groups = "drop"
  )

# Visualise the relationship between richness and Simpson diversity
ggplot(Rich_Even, aes(x = simpson, y = sobs_richness)) +
  geom_point(aes(colour = `Hospitalized?`))

```

```{r}
#| label: Evenness vs DVG Abundance Analysis

set.seed(38)

# Rarefy each sample to the minimum sequencing depth across all samples
spc.b.r <- rrarefy(dfDist, min(rowSums(dfDist)))

# Remove features that are absent after rarefaction
spc.b.r <- spc.b.r[, colSums(spc.b.r) > 0]

# Restore SRR accession numbers as row names
rownames(spc.b.r) <- H3N2f_meta$Row.names

# Convert the rarefied feature matrix to long format
rare.long <- spc.b.r %>% 
  as_tibble(rownames = "SRR") %>% 
  mutate(across(-SRR, ~ na_if(., 0))) %>% 
  pivot_longer(
    cols = -SRR,
    names_to = "Bins",
    values_to = "TotalRC",
    values_drop_na = TRUE
  )

# Combine rarefied counts with sample metadata and abundance measurements
combined_df <- list(rare.long, Meta_Data, abund) %>% 
  reduce(left_join, by = "SRR")

# Calculate Simpson diversity for each sample
ae.df <- combined_df %>% 
  group_by(SRR) %>% 
  summarise(
    simpson = diversity(TotalRC, index = "simpson")
  )

# Plot Simpson diversity against PB1 DVG abundance, coloured by hospitalisation status
merge(ae.df, combined_df, by = "SRR") %>% 
  ggplot(aes(
    x = PB1_DVG_abundance_focal_local_segment,
    y = simpson
  )) +
  geom_point(aes(colour = `Hospitalized?`)) +
  stat_ellipse(aes(colour = `Hospitalized?`))

```

## Principal Component Analysis of DVG Abundance, Coloured by Hospitalisation Status

To further characterise the relationship between DVG abundance profiles and hospitalisation status, I performed principal component analysis (PCA) to reduce the dimensionality of abundance data across all eight genomic segments for each sample.

This analysis was intended to assess whether samples clustered by hospitalisation status on the basis of their DVG abundance profiles, and by extension, whether particular DVGs were differentially abundant between hospitalised and non-hospitalised individuals. More broadly, this aimed to establish whether specific DVG community compositions are associated with hospitalisation outcome.

As this analysis is part of ongoing research, results are not reported here; the code used is provided below.

```{r}
#| label: Data Processing

#Data transformation
Merged <- merge(Depth, Meta_Data, by.x = "SRR", by.y = "Run") %>% 
  filter(!`Hospitalized?` == "Unknown") 

Merged_mat <- Merged %>%  
  pivot_longer(cols = HA_depth:PB2_depth, names_to = "Segments", values_to = "Read_Depth") %>% 
  select(SRR, Read_Depth, Segments) %>% 
  pivot_wider(names_from = SRR, values_from= Read_Depth) %>%
  as.data.frame()


rownames(Merged_mat) <- Merged_mat$Segments
Merged_mat <- Merged_mat[,-1]

#Normalize for total read depth per sample (Convert to Counts Per Million)
cpm_merged <- apply(Merged_mat, 2, function(x) (x / sum(x)) * 1000000)

#Log-transform to handle the massive range between high and low depths and adding a small pseudo-count (+1) to avoid log(0) errors
log_cpm <- log2(cpm_merged + 1)

#Running and visusalising PCA
pca_result <- prcomp(t(log_cpm), scale. = TRUE) 
p1 <- fviz_pca_ind(pca_result, repel = TRUE, addEllipses = FALSE, title = "PCA of Read Depth For Each DVG Segment", geom = "point")

#Extract the coordinates (stored in the '$x' matrix) and convert to a dataframe
pca_coords <- as.data.frame(pca_result$x)

#Adding categorical column back to these coordinates
pca_coords$Hospitalized <- Merged$`Hospitalized?`

ggplot(pca_coords, aes(x = PC1, y = PC2, color = Hospitalized)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "PCA on Segment Read Depth between Samples, Coloured by Hospitalization")+
  theme_minimal()
```

```{r}
#| label: Prepare DVG abundance matrix for PCA

# Merge DVG abundance data with sample metadata and remove samples with unknown hospitalisation status
abund_hosp <- merge(abund, Meta_Data, by.x = "SRR", by.y = "Run") %>%
  filter(`Hospitalized?` != "Unknown")

# Rename abundance columns to shorter segment names and reshape the data
ah_long <- abund_hosp %>%
  pivot_longer(
    cols = Random:PB2,
    names_to = "Segments",
    values_to = "Read_Depth"
  ) %>%
  select(SRR, Read_Depth, Segments) %>%
  pivot_wider(
    names_from = SRR,
    values_from = Read_Depth
  ) %>%
  as.data.frame()

# Use segment names as row names
rownames(ah_long) <- ah_long$Segments

# Remove the segment name column
ah_mat <- ah_long[, -1]

# Transpose the matrix so rows represent samples and columns represent segments
t_ah_mat <- t(ah_mat) %>%
  as.data.frame()

```

```{r}
#| label: Prepare relative segment abundances

# Remove the overall DVG abundance column
t_ah_mat_df <- t_ah_mat[, -1]

# Convert each sample to relative segment abundances
t_ah_mat_df <- apply(
  t_ah_mat_df,
  1,
  function(x) x / sum(x)
) %>%
  t()

# Inspect samples with unexpected values
t_ah_mat_df[c("SRR7889156", "SRR7889183"), ]

# Replace rows containing only missing values with zeros
t_ah_mat_df[is.na(rowSums(t_ah_mat_df)), ] <- 0
```

```{r}
#| label: Perform PCA and Visualise Results

# Run PCA on relative segment abundances
pca_ah <- prcomp(t_ah_mat_df, scale. = TRUE)

# Extract sample coordinates in principal component space
pca_coords <- as.data.frame(pca_ah$x)

# Add sample metadata for plotting
pca_coords$Hospitalized <- abund_hosp$`Hospitalized?`

# Classify samples by overall DVG abundance
pca_coords <- pca_coords %>%
  mutate(
    label = ifelse(
      t_ah_mat$Random > 0.5,
      "Over 50",
      "Under 50"
    )
  )

# Create subsets for high- and low-abundance samples
over_50 <- t_ah_mat %>%
  filter(Random > 0.5)

under_50 <- t_ah_mat %>%
  filter(Random <= 0.5)

# Display the variance explained by each principal component
screeplot(
  pca_ah,
  type = "lines",
  main = "Scree Plot"
)

# PCA scatter plot using ggplot2
ggplot(
  pca_coords,
  aes(x = PC1, y = PC2, colour = label)
) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse() +
  geom_quasirandom(
    method = "smiley",
    varwidth = TRUE,
    size = 3
  ) +
  labs(
    title = "PCA of Relative DVG Segment Abundance",
    colour = "Overall DVG abundance"
  ) +
  theme_minimal()

# PCA scatter plot with confidence ellipses using factoextra
fviz_pca_ind(
  pca_ah,
  repel = TRUE,
  addEllipses = TRUE,
  title = "PCA of Relative DVG Segment Abundance",
  habillage = pca_coords$label,
  ellipse.type = "confidence",
  geom = "point",
  pointsize = 3
)

# Prepare the first three principal components for 3D visualisation
data.frame(
  PC1 = pca_ah$x[, 1],
  PC2 = pca_ah$x[, 2],
  PC3 = pca_ah$x[, 3],
  Group = t_ah_mat$Random
)

# Create an interactive 3D PCA plot
plot_ly(
  data = threedim_coordinates,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~Group,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 5)
) %>%
  layout(
    scene = list(
      xaxis = list(title = "PC1"),
      yaxis = list(title = "PC2"),
      zaxis = list(title = "PC3")
    )
  )

```

```{r}
#| label: DVG abundance of polymerase segments vs DVG abundance of random

polvall <- abund |> 
  mutate(pol_abund = PB1 + PB2 + PA) |> 
  mutate(totalDVG = rowSums(across(3:10), na.rm = TRUE)) |> 
  mutate(percentage_pol = pol_abund / totalDVG)

ggplot(polvall, aes(x = Random, y = percentage_pol)) + 
  geom_point()+
  theme_minimal()
```
